Categories of Optics (Draft)

### 2018-08-18 paper.tex
Fix up lawful profunctor bit

\newcommand{\fget}{\textsc{Get}}
\newcommand{\fput}{\textsc{Put}}
\newcommand{\fmodify}{\textsc{Modify}}
\newcommand{\freview}{\textsc{Review}}
\newcommand{\fcreate}{\textsc{Create}}
\newcommand{\fmatching}{\textsc{Matching}}
\newcommand{\funzip}{\textsc{Unzip}}
\newcommand{\fover}{\textsc{Over}}

@@

\title{Categories of Optics}
\author{Mitchell Riley}
\affil{Wesleyan University \\ \texttt{mvriley@wesleyan.edu}}
\begin{document}
\maketitle

@@

So that our lenses better conform to this intuitive idea of ``zooming in'', we often want them satisfy some conditions known as the \emph{lens laws}:
\begin{center}
\begin{minipage}[b]{0.33333\textwidth}
\begin{center}
\[
\begin{tikzcd}
S \times A \ar[rr, "\fput"] \ar[dr, "\pi_2", swap] && S \ar[dl, "\fget"] \\
& A
\end{tikzcd}
\]
\hspace{0.8cm}$\fput\fget$
\end{center}
\end{minipage}%
\begin{minipage}[b]{0.33333\textwidth}
\begin{center}
\[
\begin{tikzcd}
S \ar[rr, "{[\id_S, \fget]}"] \ar[dr, "\id_S", swap] && S \times A \ar[dl, "\fput"] \\
& S
\end{tikzcd}
\]
\hspace{-0.6cm}$\fget\fput$
\end{center}
\end{minipage}%
\begin{minipage}[b]{0.33333\textwidth}
\begin{center}
\[
\begin{tikzcd}
S \times A \times A \ar[r, "\fput \times A"] \ar[d, "\pi_{1, 3}", swap] & S \times A\ar[d, "\fput"] \\
S \times A \ar[r, "\fput", swap] & A
\end{tikzcd}
\]
\quad$\fput\fput$
\end{center}
\end{minipage}%
\end{center}
We call such lenses lawful. The $\fput\fget$ law states that any update to $A$ is represented faithfully in $S$. The $\fget\fput$ law states that if $A$ is not changed then neither is $S$; and finally, the $\fput\fput$ law states that any update to $A$ completely overwrites previous updates.

@@

Lenses were discovered to be just one of a hierarchy of data accessors, including prisms, setters, traversals and more. These are collectively called \emph{optics} and have been best explored in the widely uesd Haskell \lenslib{} library: see~\cite{LensLibrary}. Each optic variant has a concrete description as a certain collection of maps, with attendant laws under which we consider them well-behaved. We begin in Section~\ref{sec:optics} by defining the \emph{category of optics} in a sufficiently general way to encompass almost all the optic variants in use in the wild, using lenses as a running example. The category of lenses is precisely the result of this construction when applied to a category with finite products. Section~\ref{sec:lawful-optics} defines the equivalent of the lens laws for a general category of optics. Then in Section~\ref{sec:examples} we see that these generic definitions specialise correctly to the other basic varieties of optic, including the laws. We include in Section~\ref{sec:mixed-optics} a more speculative discussion of ``mixed optics'', which include indexed and coindexed lenses and traversals.

In the Haskell \lenslib{} library, optics are written in a form known as the \emph{profunctor encoding}, which at first glance is completely different to that given in Section~\ref{sec:optics}. As this was being written, Bartosz Milewski~\cite{ProfunctorOpticsPost} independently described the isomorphism between some forms of optic and their profunctor encoding. In Section~\ref{sec:profunctor-optics} we extend this to more of the optic hierarchy and translate the laws into the profunctor setting.

@@

Much of what is known about lenses and their variants is folklore, and careful verification of some of their categorical properties has been lacking. The aim of the present paper is to fill this gap, with the hope that a better understanding of the general structure of these categories will make it easier to generalise to new and exotic settings. This is particularly important with the advent of linear types in Haskell, which enables a new branch of the lens family tree, and also with the new applications to game theory.

\subsection{Contributions}
\todo{
  \begin{itemize}
  \item A definition of optics that uses the monoidal action of one category of another. This general enough to encompass lenses, setters, traversals, and more exotic things.
  \item A definition of lawfulness that specialises in the correct way in and allows us to easily derive concrete laws for new kids of optic. Importantly, in the general case this definition is less strict than the `constant complement' definition.
  \item A universal property satisfied by the category of optics in the monoidal case
  \item A categorical description of optic laws as the homomorphism condition on a morphism between certain comonoids
  \item An attempt to isolate what $\Set$-like properties are being used in some places
  \end{itemize}
}

@@

\begin{lemma}[Coend as coequaliser]\label{lemma:calculate-coend}
If $\E$ is cocomplete, the coend of $P : \M^\op \times \M \to \E$ can be calculated as the coequaliser in the diagram
\[
  \begin{tikzcd}
    \displaystyle \coprod_{M \to N} P(N, M) \ar[r,shift left=.75ex]  \ar[r,shift right=.75ex] & \displaystyle\coprod_{M \in \M} P(M, M) \ar[r] & \displaystyle\int^{M \in \M} P(M, M)
  \end{tikzcd}
\]
\qed
\end{lemma}

@@

  The residual $M$ should be thought of as a kind of `scratch space'; information from $S$ that we need to remember to construct $S'$. The quotienting imposed by the coend means we cannot inspect this temporary information, indeed, there is not even a canonical choice for the object $M$ in general, given an optic $S \hto A$.

  Elements of $\Optic_\C((S, S'), (A, A'))$ have an appealing
  interpretation as string diagrams with a ``hole'' missing. We draw the
  pair $\rep{l}{r}$ as
  \begin{center}
    \input{diagrams/generic-optic.tikz}
  \end{center}
  reading left to right, so the left half of the diagram represents $l$ and the right half $r$. The relation expressed by the coend can be graphically expressed as:
  \begin{center}
    \input{diagrams/coend-relation-left.tikz}
    \hspace{0.7cm} \raisebox{1.35cm}{$\sim$} \hspace{1cm}
    \input{diagrams/coend-relation-right.tikz}
  \end{center}
  We will therefore omit the vertical cut between $l$ and $r$ in most subsequent diagrams; any choice yields a representative of same optic.

@@

\begin{theorem}\label{thm:optic-is-free-teleological-cat}
  $\Optic : \StrictSymmMonCat \to \StrictTele$ is left adjoint to the `underlying dualisable morphisms' functor ${(-)}_d : \StrictTele \to \StrictSymmMonCat$.
\end{theorem}
\begin{proof}
Precomposition with $j$ gives a function
\begin{align*}
\StrictTele(\Optic_\C, \T) \to \StrictSymmMonCat(\C, \T_d)
\end{align*}
and the previous proposition states that this is a isomorphism.

This is automatically natural in $\T$. Naturality in $\C$ follows by Lemma~\ref{lem:iota-commute-with-opticf}. 
\end{proof}

@@

\begin{remark}
We use $;$ to denote composition of $\C$ in diagrammatic order. The reason for this is that the coend relation can be applied simply by shifting the position of $\mid$ in a representative:
\begin{align*}
\rep{l;\phi A}{r} = \rep{l}{\phi A;r}
\end{align*}

\todo{Should I switch the entire paper into using diagrammatic order?}
\end{remark}

Let $\Twoptic_\M(S, A)$ \todo{(please come up with a better notation)} denote the set \[ \int^{M_1, M_2 \in \M} \C(S, M_1 A) \times \C(M_1 A, M_2 A) \times \C(M_2 A, S). \]
Using the universal property of the coend, we define three maps:
\begin{align*}
  \outside &: \Optic_\M(S, A) \to \C(S, S) \\
  \once, \twice &: \Optic_\M(S, A) \to \Twoptic_\M(S, A)
\end{align*}
by
\begin{align*}
  \outside(\rep{l}{r}) &= l;r \\
  \once(\rep{l}{r}) &= \repthree{l}{\id_{MA}}{r} \\
  \twice(\rep{l}{r}) &= \repthree{l}{r;l}{r}
\end{align*}

@@

  We now have to slog through evaluating $\Phi(\once(p))$ and $\Phi(\twice(p))$.
  \begin{alignat*}{3}
    \Phi(\once(\rep{[\id_S, \fget]}{\fput})) &=
    \Phi(&&\repthree{[\id_S, \fget]}{\id_{S \times A}}{\fput}) \\
    &= (&&\pi_2 [\id_S, \fget], \\
    &&&\pi_2 \id_{S \times A} (\pi_1 [\id_S, \fget] \times A), \\
    &&& \fput (\pi_1 \id_{S \times A} (\pi_1 [\id_S, \fget] \times A) \times A) \quad) \\
    %
    &= (&&\fget, \\
    &&& \pi_2, \\
    &&& \fput \pi_{1,3} \quad) \\
    %%%%
    %%%%
    \Phi(\twice(\rep{[\id_S, \fget]}{\fput})) &=
    \Phi(&&\repthree{[\id_S, \fget]}{[\id_S, \fget]\fput}{\fput}) \\
    &= (&&\pi_2 [\id_S, \fget], \\
    &&&\pi_2 [\id_S, \fget]\fput (\pi_1 [\id_S, \fget] \times A), \\
    &&& \fput (\pi_1 [\id_S, \fget]\fput (\pi_1 [\id_S, \fget] \times A) \times A) \quad) \\
    %
    &= (&&\fget, \\
    &&& \fget \; \fput (\id_S \times A), \\
    &&& \fput (\fput (\id_S \times A) \times A) \quad) \\
    %
    &= (&&\fget, \\
    &&& \fget \; \fput, \\
    &&& \fput (\fput \times A) \quad)
  \end{alignat*}
  So checking component-wise, $\Phi(\once(p))$ being equal to $\Phi(\twice(p))$ is exactly equivalent to the $\fput\fget$ and $\fput\fput$ laws holding.
\end{proof}

@@

\begin{remark}
  In the case that $\C$ is cartesian closed, $\Optic_\times$ is monoidal closed via the astonishing formula
  \begin{align*}
    [(S, S'), (A, A')] := (\homC(S, A) \times \homC(S \times A', S'), \, S \times A')
  \end{align*}
  where $\homC(-, -)$ denotes the internal hom. For a proof see~\cite[Section 1.2]{DialecticaCategories}. This cannot be extended to non-cartesian closed categories, the isomorphism
  \begin{align*}
    \Optic_\times((S, S') \otimes (T, T'), (A, A')) \cong \Optic_\times((S, S'),  [(T, T'), (A, A')])
  \end{align*}
  uses the diagonal maps of $\C$ in an essential way.
\end{remark}

@@

  Now take $l : [j,\fget] : S \to C \times A$  and $r : \fput (i \times A) : C \times A \to S$. That they are mutual inverses is easily checked:
  \begin{align*}
    \fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
                                &= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
                                &= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
                                &= \id_S && \text{(by $\fget\fput$)}
  \intertext{and}
    [j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
                                &= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
                                &= [j\fput (i \times A), \pi_2] && \\
                                &= [jij\fput (i \times A), \pi_2] && \\
                                &= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
                                &= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
                                &= [jij \pi_1 (i \times A), \pi_2] && \\
                                &= [jiji \pi_1, \pi_2] && \\
                                &= [\pi_1, \pi_2] && \\
                                &= \id_{C \times A}
  \end{align*}

@@

\begin{definition}
  Suppose $\C$ has finite coproducts. The \emph{category of prisms} is the category of lawful optics with respect to the coproduct $\sqcup$.
\end{definition}

Just as optics for $\times$ correspond to a pair of maps $\fget : S \to A$ and $\fput : S \times A \to S$, optics for $\sqcup$ correspond to pairs of maps $\freview : A \to S$ and $\fmatching : S \to S \sqcup A$. These names are taken from the Haskell \lenslib{} library.
\begin{align*}
  \Optic_\sqcup((S, S'), (A, A')) &= \int^{M \in \C} \C(S, M \sqcup A) \times \C(M \sqcup A', S') \\
                                  &\cong \int^{M \in \C} \C(S, M \sqcup A) \times \C(M, S') \times \C(A, S') && \text{(universal property of coproduct)} \\
                                  &\cong \C(S, S' \sqcup A) \times \C(A', S') && \text{(Yoneda reduction)}
\end{align*}
If we are given a prism $\rep{l}{r} : (S, S') \hto (A, A')$ then associated $\freview$ and $\fmatching$ morphisms are given by $\freview = r \inr$ and $\fmatching = (r\inl \sqcup A)l$

@@

There is a common pattern in many of the examples to follow: for every object $A \in \C$, the evaluation-at-$A$ functor $- \act A : \M \to \C$ has a right adjoint, say $R_A : \C \to \M$. To fix notation, let \[\overrightarrow{(-)} : \C(F \act A, S) \to \M(F, R_{A} S) : \overleftarrow{(-)}\] denote the homset bijection, so the unit and counit are:
    \begin{equation*}
    \begin{aligned}[t]
  	\eta_F &: F \to R_A (F \act A) \\
    \eta_F &:= \overrightarrow{\id_{F \act A}}
    \end{aligned}
    \qquad\qquad\qquad
    \begin{aligned}[t]
    \varepsilon_{S} &: (R_A S) \cdot A \to S \\
    \varepsilon_{S} &:= \overleftarrow{\id_{R_{A} S}}
    \end{aligned}
\end{equation*}
It is shown in~\cite[Section 6]{ANoteOnActions} that, at least in the case $\M$ is right closed, to give such an action is equivalent to giving a ``copowered $\M$-category $\C$''. In most cases of interest to us, however, $\M$ is not right closed. \todo{Pointed out to me years ago when I emailed Ross Street, but I didn't understand what he was saying}

In this case we can always eliminate the coend and find a concrete description of an optic:
\begin{align*}
&\int^{F \in \M} \C(S, F\act A) \times \C(F\act A', S') \\
&\cong \int^{F \in \M} \C(S, F \act A) \times \M(F, R_{A'} S') \\
&\cong \C(S,  (R_{A'} S') \act A)
\end{align*}
A concrete optic is therefore a map $\funzip : S \to (R_{A'} S') \act A$. The above isomorphism sends $\funzip$ to the element $\rep{\funzip}{\varepsilon_{S'}}$. In the other direction, given $\rep{l}{r}$ we have $\funzip = (\overrightarrow{r} \act A)l$.

@@

So now interpreting the optic laws, find
\[\outside(\rep{\funzip}{\varepsilon_{S} }) = \varepsilon_{S} \; \funzip = \id_S \] is exactly the coalgebra unit law, and equality of
  \begin{align*}
    \Phi(\once(\rep{\funzip}{\varepsilon_S}))
    &= \Phi(\repthree{\funzip}{\id_{(R_{A} S) \act A}}{\varepsilon_S }) \\
    &= (\overrightarrow{(\overrightarrow{\varepsilon_S} \act A)\id_{(R_{A} S) \act A}} \act A)\funzip \\
    &= (\overrightarrow{(\id_{R_A S} \act A)} \act A)\funzip \\
    &= (\overrightarrow{\id_{R_A S \act A}} \act A)\funzip \\
    &= (\eta_{R_A S} \act A) \funzip \\
    \Phi(\twice(\rep{\funzip}{\varepsilon_S }))
    &= \Phi(\repthree{\funzip}{\funzip \; \varepsilon_S}{\varepsilon_S }) \\
    &= (\overrightarrow{(\overrightarrow{\varepsilon_S} \act A)(\funzip \; \varepsilon_S)} \act A)\funzip \\
    &= (\overrightarrow{(\funzip \; \varepsilon_S)} \act A)\funzip \\
    &= (R_A (\funzip) \act A)\funzip
  \end{align*}
is exactly the coalgebra multiplication law.
\end{proof}

@@

We could immediately use the previous section to give a coalgebraic description of setters and their laws, but with a little more manipulation we get a form that looks more familiar:
\begin{align*}
  \Setter_\C((S, S'), (A, A')) &= \int^{F \in [\C, \C]} \C(S, FA) \times \C(FA', S') \\
                               &\cong \int^{F \in [\C, \C]} [\C, \C](\C(A, -) \cdot S, F) \times \C(FA', S') \\
                               &\cong \C(\C(A, A') \cdot S, S') \\
                               &\cong \Set(\C(A, A'), \C(S, S'))
\end{align*}

@@

\begin{proposition}
A setter $p : S \hto A$ is lawful iff
\begin{align*}
\fover(\id_A) &= \id_S \\
\fover(f)\fover(g)&= \fover(fg)
\end{align*}
\end{proposition}
\begin{proof}
The key is concretely describing $\Twoptic_{[\C, \C]}(S, A)$ as \[ \Set( \C(A, A) \times \C(A, A), \C(S, S) ).\] We leave the verification that the conditions are equivalent to the reader. \todo{Worth actually writing this out?}
%Because a concrete setter is a function in $\Set$, it is enough to verify the equations pointwise.
\end{proof}

This characterisation of setters is maybe a little odd, in that we have ended up with a function of $\Set$s, rather than a description internal to $\C$. If we modify our definition of $\Setter$, we do get an internal characterisation. Suppose $\C$ is cartesian closed and let $\Strong_\C$ be the category of \emph{strong functors} on $\C$. \todo{I define strong functor here because I also want to talk about strong monads later, should this definition wait for that?}

@@

\begin{definition}
The category $\Traversal$ of traversals is the category of lawful optics for the action of $\Traversable$ on $\Set$ given by evaluation. (Yes, the names $\Traversal$ vs\@. $\Traversable$ are confusing!)
\end{definition}

>>>>

\newcommand{\fget}{\textsc{Get}}
\newcommand{\fput}{\textsc{Put}}
\newcommand{\fmodify}{\textsc{Modify}}
\newcommand{\freview}{\textsc{Review}}
\newcommand{\fcreate}{\textsc{Create}}
\newcommand{\fmatching}{\textsc{Matching}}
\newcommand{\funzip}{\textsc{Unzip}}
\newcommand{\fover}{\textsc{Over}}
\newcommand{\findex}{\textsc{Index}}

@@

\title{Categories of Optics (Draft)}
\author{Mitchell Riley}
\affil{Wesleyan University \\ \texttt{mvriley@wesleyan.edu}}
\begin{document}
\maketitle

@@

So that our lenses better adhere to this intuitive idea of ``zooming in'', we often want them satisfy some conditions known as the \emph{lens laws}:
\begin{center}
\begin{minipage}[b]{0.33333\textwidth}
\begin{center}
\[
\begin{tikzcd}
S \times A \ar[rr, "\fput"] \ar[dr, "\pi_2", swap] && S \ar[dl, "\fget"] \\
& A
\end{tikzcd}
\]
\hspace{0.8cm}$\fput\fget$
\end{center}
\end{minipage}%
\begin{minipage}[b]{0.33333\textwidth}
\begin{center}
\[
\begin{tikzcd}
S \ar[rr, "{[\id_S, \fget]}"] \ar[dr, "\id_S", swap] && S \times A \ar[dl, "\fput"] \\
& S
\end{tikzcd}
\]
\hspace{-0.6cm}$\fget\fput$
\end{center}
\end{minipage}%
\begin{minipage}[b]{0.33333\textwidth}
\begin{center}
\[
\begin{tikzcd}
S \times A \times A \ar[r, "\fput \times A"] \ar[d, "\pi_{1, 3}", swap] & S \times A\ar[d, "\fput"] \\
S \times A \ar[r, "\fput", swap] & A
\end{tikzcd}
\]
\quad$\fput\fput$
\end{center}
\end{minipage}%
\end{center}
We call such lenses lawful. The $\fput\fget$ law states that any update to $A$ is represented faithfully in $S$. The $\fget\fput$ law states that if $A$ is not changed then neither is $S$; and finally, the $\fput\fput$ law states that any update to $A$ completely overwrites previous updates.

@@

Lenses were discovered to be just one of a hierarchy of data accessors, including prisms, setters, traversals and more. These are collectively called \emph{optics} and have been best explored in the widely uesd Haskell \lenslib{} library: see~\cite{LensLibrary}. Each optic variant has a concrete description as a certain collection of maps, with attendant laws under which we consider them well-behaved. We begin in Section~\ref{sec:optics} by defining the \emph{category of optics} in a sufficiently general way to encompass almost all the optic variants in use in the wild, using lenses as a running example. The category of lenses is precisely the result of this construction when applied to a category with finite products. Section~\ref{sec:lawful-optics} defines the equivalent of the lens laws for a general category of optics. Then in Section~\ref{sec:examples} we see that these generic definitions specialise correctly to the other basic varieties of optic, including the laws. % We include in Section~\ref{sec:mixed-optics} a more speculative discussion of ``mixed optics'', which include indexed and coindexed lenses and traversals.

In the Haskell \lenslib{} library, optics are written in a form known as the \emph{profunctor encoding}, which at first glance is completely different to that given in Section~\ref{sec:optics}. As this was being written, Bartosz Milewski~\cite{ProfunctorOpticsPost} independently described the isomorphism between some forms of optic and their profunctor encoding. In Section~\ref{sec:profunctor-optics} we review this equivalence and verify that the folklore profunctor optic laws are equivalent to lawfulness as defined here.

@@

Much of what is known about optics is folklore, and careful verification of some of their categorical properties has been lacking. The aim of the present paper is to fill this gap, with the hope that a better understanding of the general structure of these categories will make it easier to generalise to new and exotic settings. This is particularly important with the advent of linear types in Haskell, enabling a new branch of the lens family tree, and also with the new applications to game theory.

\subsection{Contributions}
  \begin{itemize}
  \item An careful account of the folkloric optic construction in an arbitrary symmetric monoidal category $\C$, summarised compactly as giving a functor $\Optic : \SymmMonCat \to \SymmMonCat$ (Section~\ref{sec:optics}),
  \item A characterisation of the $\Optic$ construction as freely adding counits to a set of `dualisable morphisms' (Section~\ref{sec:teleological-categories}),
  \item A definition of lawfulness for a general optic category that specialises in the correct way to known cases and allows us to easily derive concrete laws for new kids of optic (Section~\ref{sec:lawful-optics}),
  \item Commentary on the optic variants used most frequently in the wild (Section~\ref{sec:examples}),
  \item A proof that lawfulness as defined here is equivalent to the folklore profunctor lens laws. (Section~\ref{sec:profunctor-optics})
  \end{itemize}

@@

\begin{lemma}[Coend as coequaliser]\label{lemma:calculate-coend}
If $\E$ is cocomplete and $\M$ is small, the coend of $P : \M^\op \times \M \to \E$ can be calculated as the coequaliser in the diagram
\[
  \begin{tikzcd}
    \displaystyle \coprod_{M \to N} P(N, M) \ar[r,shift left=.75ex]  \ar[r,shift right=.75ex] & \displaystyle\coprod_{M \in \M} P(M, M) \ar[r] & \displaystyle\int^{M \in \M} P(M, M)
  \end{tikzcd}
\]
\qed
\end{lemma}

@@

  The residual $M$ should be thought of as a kind of `scratch space'; information from $S$ that we need to remember to construct $S'$. The quotienting imposed by the coend means we cannot inspect this temporary information, indeed, given an optic $S \hto A$ there is not even a canonical choice for the object $M$ in general.

  Elements of $\Optic_\C((S, S'), (A, A'))$ have an appealing
  interpretation as string diagrams with a ``hole'' missing. We draw the
  pair $\rep{l}{r}$ as
  \begin{center}
    \input{diagrams/generic-optic.tikz}
  \end{center}
  reading left to right, so the portion of the diagram to the left of the line represents $l$ and the right portion $r$. The relation expressed by the coend can be drawn graphically as:
  \begin{center}
    \input{diagrams/coend-relation-left.tikz}
    \hspace{0.7cm} \raisebox{1.35cm}{$\sim$} \hspace{1cm}
    \input{diagrams/coend-relation-right.tikz}
  \end{center}
  We will therefore omit the vertical cut between $l$ and $r$ in most subsequent diagrams; any choice yields a representative of same optic.

@@

\begin{theorem}\label{thm:optic-is-free-teleological-cat}
  $\Optic : \StrictSymmMonCat \to \StrictTele$ is left adjoint to the `underlying dualisable morphisms' functor ${(-)}_d : \StrictTele \to \StrictSymmMonCat$.
\end{theorem}
\begin{proof}
Precomposition with $j$ gives a function
\begin{align*}
\StrictTele(\Optic_\C, \T) \to \StrictSymmMonCat(\C, \T_d)
\end{align*}
and the previous proposition states that this is a isomorphism. This is automatically natural in $\T$. Naturality in $\C$ follows by Lemma~\ref{lem:iota-commute-with-opticf}. 
\end{proof}

@@

\begin{remark}
We use $;$ to denote composition of $\C$ in diagrammatic order. The reason for this is that the coend relation can be applied simply by shifting the position of $\mid$ in a representative:
\begin{align*}
\rep{l;\phi A}{r} = \rep{l}{\phi A;r}
\end{align*}
%\todo{Should I switch the entire paper into using diagrammatic order?}
\end{remark}

Let $\Twoptic_\M(S, A)$ denote the set \[ \int^{M_1, M_2 \in \M} \C(S, M_1 A) \times \C(M_1 A, M_2 A) \times \C(M_2 A, S). \]
Using the universal property of the coend, we define three maps:
\begin{align*}
  \outside &: \Optic_\M(S, A) \to \C(S, S) \\
  \once, \twice &: \Optic_\M(S, A) \to \Twoptic_\M(S, A)
\end{align*}
by
\begin{align*}
  \outside(\rep{l}{r}) &= l;r \\
  \once(\rep{l}{r}) &= \repthree{l}{\id_{MA}}{r} \\
  \twice(\rep{l}{r}) &= \repthree{l}{r;l}{r}
\end{align*}

@@

  We now have to slog through evaluating $\Phi(\once(p))$ and $\Phi(\twice(p))$.
  \begingroup
  \allowdisplaybreaks
  \begin{alignat*}{3}
    \Phi(\once(\rep{[\id_S, \fget]}{\fput})) &=
    \Phi(&&\repthree{[\id_S, \fget]}{\id_{S \times A}}{\fput}) \\
    &= (&&\pi_2 [\id_S, \fget], \\
    &&&\pi_2 \id_{S \times A} (\pi_1 [\id_S, \fget] \times A), \\
    &&& \fput (\pi_1 \id_{S \times A} (\pi_1 [\id_S, \fget] \times A) \times A) \quad) \\
    %
    &= (&&\fget, \\
    &&& \pi_2, \\
    &&& \fput \pi_{1,3} \quad) \\
    %%%%
    %%%%
    \Phi(\twice(\rep{[\id_S, \fget]}{\fput})) &=
    \Phi(&&\repthree{[\id_S, \fget]}{[\id_S, \fget]\fput}{\fput}) \\
    &= (&&\pi_2 [\id_S, \fget], \\
    &&&\pi_2 [\id_S, \fget]\fput (\pi_1 [\id_S, \fget] \times A), \\
    &&& \fput (\pi_1 [\id_S, \fget]\fput (\pi_1 [\id_S, \fget] \times A) \times A) \quad) \\
    %
    &= (&&\fget, \\
    &&& \fget \; \fput (\id_S \times A), \\
    &&& \fput (\fput (\id_S \times A) \times A) \quad) \\
    %
    &= (&&\fget, \\
    &&& \fget \; \fput, \\
    &&& \fput (\fput \times A) \quad)
  \end{alignat*}
  \endgroup
  So checking component-wise, $\Phi(\once(p))$ being equal to $\Phi(\twice(p))$ is exactly equivalent to the $\fput\fget$ and $\fput\fput$ laws holding.
\end{proof}

@@

\begin{remark}
  In the case that $\C$ is cartesian closed, $\Optic_\times$ is monoidal closed via the astonishing formula
  \begin{align*}
    [(S, S'), (A, A')] := (\homC(S, A) \times \homC(S \times A', S'), \, S \times A')
  \end{align*}
  where $\homC(-, -)$ denotes the internal hom. For a proof see~\cite[Section 1.2]{DialecticaCategories}. This cannot be extended to non-cartesian closed categories, the isomorphism
  \begin{align*}
    \Lens((S, S') \otimes (T, T'), (A, A')) \cong \Lens((S, S'),  [(T, T'), (A, A')])
  \end{align*}
  uses the diagonal maps of $\C$ in an essential way.
\end{remark}

@@

  Now take $l : [j,\fget] : S \to C \times A$  and $r : \fput (i \times A) : C \times A \to S$. That they are mutual inverses is easily checked:
  \begin{align*}
    \fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
                                &= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
                                &= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
                                &= \id_S && \text{(by $\fget\fput$)}
  \intertext{and}
    [j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
                                &= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
                                &= [j\fput (i \times A), \pi_2] && \\
                                &= [jij\fput (i \times A), \pi_2] && \\
                                &= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
                                &= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$)}\\ %  \todo{todo: detail}
                                &= [jij \pi_1 (i \times A), \pi_2] && \\
                                &= [jiji \pi_1, \pi_2] && \\
                                &= [\pi_1, \pi_2] && \\
                                &= \id_{C \times A}
  \end{align*}

@@

\begin{definition}
  Suppose $\C$ has finite coproducts. The \emph{category of prisms} is the category of lawful optics with respect to the coproduct $\sqcup$: $\Prism \defeq \Optic_\sqcup$.
\end{definition}

Just as optics for $\times$ correspond to a pair of maps $\fget : S \to A$ and $\fput : S \times A \to S$, optics for $\sqcup$ correspond to pairs of maps $\freview : A \to S$ and $\fmatching : S \to S \sqcup A$. These names are taken from the Haskell \lenslib{} library.
\begin{align*}
  \Prism((S, S'), (A, A')) &= \int^{M \in \C} \C(S, M \sqcup A) \times \C(M \sqcup A', S') \\
                                  &\cong \int^{M \in \C} \C(S, M \sqcup A) \times \C(M, S') \times \C(A, S') && \text{(universal property of coproduct)} \\
                                  &\cong \C(S, S' \sqcup A) \times \C(A', S') && \text{(Yoneda reduction)}
\end{align*}
If we are given a prism $\rep{l}{r} : (S, S') \hto (A, A')$ then associated $\freview$ and $\fmatching$ morphisms are given by $\freview = r \inr$ and $\fmatching = (r\inl \sqcup A)l$

@@

There is a common pattern in many of the examples to follow: for every object $A \in \C$, the evaluation-at-$A$ functor $- \act A : \M \to \C$ has a right adjoint, say $R_A : \C \to \M$. To fix notation, let \[\overrightarrow{(-)} : \C(F \act A, S) \to \M(F, R_{A} S) : \overleftarrow{(-)}\] denote the homset bijection, so the unit and counit are:
    \begin{equation*}
    \begin{aligned}[t]
  	\eta_F &: F \to R_A (F \act A) \\
    \eta_F &:= \overrightarrow{\id_{F \act A}}
    \end{aligned}
    \qquad\qquad\qquad
    \begin{aligned}[t]
    \varepsilon_{S} &: (R_A S) \cdot A \to S \\
    \varepsilon_{S} &:= \overleftarrow{\id_{R_{A} S}}
    \end{aligned}
\end{equation*}
It is shown in~\cite[Section 6]{ANoteOnActions} that, at least in the case $\M$ is right closed, to give such an action is equivalent to giving a ``copowered $\M$-category $\C$''. In most cases of interest to us, however, $\M$ is not right closed. %\todo{Pointed out to me years ago when I emailed Ross Street, but I didn't understand what he was saying}

When we have such a right adjoint, we can always find a concrete description of an optic:
\begin{align*}
\Optic_\M((S, S'), (A, A')) &= \int^{F \in \M} \C(S, F\act A) \times \C(F\act A', S') \\
&\cong \int^{F \in \M} \C(S, F \act A) \times \M(F, R_{A'} S') \\
&\cong \C(S,  (R_{A'} S') \act A)
\end{align*}
A concrete optic is therefore a map $\funzip : S \to (R_{A'} S') \act A$. The above isomorphism sends $\funzip$ to the element $\rep{\funzip}{\varepsilon_{S'}}$. In the other direction, given $\rep{l}{r}$ we have $\funzip = (\overrightarrow{r} \act A)l$.

@@

So now interpreting the optic laws, we find
\[\outside(\rep{\funzip}{\varepsilon_{S} }) = \varepsilon_{S} \; \funzip = \id_S \] is exactly the coalgebra unit law, and equality of
  \begin{align*}
    \Phi(\once(\rep{\funzip}{\varepsilon_S}))
    &= \Phi(\repthree{\funzip}{\id_{(R_{A} S) \act A}}{\varepsilon_S }) \\
    &= (\overrightarrow{(\overrightarrow{\varepsilon_S} \act A)\id_{(R_{A} S) \act A}} \act A)\funzip \\
    &= (\overrightarrow{(\id_{R_A S} \act A)} \act A)\funzip \\
    &= (\overrightarrow{\id_{R_A S \act A}} \act A)\funzip \\
    &= (\eta_{R_A S} \act A) \funzip \\
    \Phi(\twice(\rep{\funzip}{\varepsilon_S }))
    &= \Phi(\repthree{\funzip}{\funzip \; \varepsilon_S}{\varepsilon_S }) \\
    &= (\overrightarrow{(\overrightarrow{\varepsilon_S} \act A)(\funzip \; \varepsilon_S)} \act A)\funzip \\
    &= (\overrightarrow{(\funzip \; \varepsilon_S)} \act A)\funzip \\
    &= (R_A (\funzip) \act A)\funzip
  \end{align*}
is exactly the coalgebra multiplication law.
\end{proof}

@@

We could immediately use the previous section to give a coalgebraic description of setters and their laws, but with a little manipulation we get a form that looks more familiar:
\begin{align*}
  \Setter_\C((S, S'), (A, A')) &= \int^{F \in [\C, \C]} \C(S, FA) \times \C(FA', S') \\
                               &\cong \int^{F \in [\C, \C]} [\C, \C](\C(A, -) \cdot S, F) \times \C(FA', S') \\
                               &\cong \C(\C(A, A') \cdot S, S') \\
                               &\cong \Set(\C(A, A'), \C(S, S'))
\end{align*}

@@

\begin{proposition}
A setter $p : S \hto A$ is lawful iff
\begin{align*}
\fover(\id_A) &= \id_S \\
\fover(f)\fover(g)&= \fover(fg)
\end{align*}
\end{proposition}
\begin{proof}
The key is concretely describing $\Twoptic_{[\C, \C]}(S, A)$ as \[ \Set( \C(A, A) \times \C(A, A), \C(S, S) ).\] We leave the verification that the conditions are equivalent to the reader. %\todo{Worth actually writing this out?}
%Because a concrete setter is a function in $\Set$, it is enough to verify the equations pointwise.
\end{proof}

This characterisation of setters is maybe a little odd, in that we have ended up with a function of $\Set$s, rather than a description internal to $\C$. If we modify our definition of $\Setter$, we do get an internal characterisation. Suppose $\C$ is cartesian closed and let $\Strong_\C$ be the category of \emph{strong functors} on $\C$. %\todo{I define strong functor here because I also want to talk about strong monads later, should this definition wait for that?}

@@

\begin{definition}
The category $\Traversal$ of traversals is the category of lawful optics for the action of $\Traversable$ on $\Set$ given by evaluation. (Yes, the names $\Traversal$/$\Traversable$ are confusing!)
\end{definition}
